flowCore: a Bioconductor package for high throughput flow cytometry

### 2009-03-18 paper/bmc_article.tex
response2reviewers and manuscript amendements
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38020 bc3139a8-67e5-0310-9ffc-ced21a209358

\title{flowCore a Bioconductor package for high throughput flow cytometry}

@@

%% I think switching the last 2 paragraphs make more sense as the 2nd
%% one of this section talk about other closed-source soft etc.
%% But up to you
\Rpackage{flowCore} does not provide a graphical user interface
(\textit{i.e.}, is not a GUI-driven software) and all operations are
done using a command line interface.  It is possible to add a more
elaborate user interface on top of this infrastructure, however the
focus in this paper is on a programmatic approach to enable the
convenient development of novel analysis methods and automation of
complex analysis approaches.  By taking the burden of data management
from the programmer, and by providing well-defined application
programming interfaces (APIs), it is possible to readily test new
ideas and to easily extend the framework's functionality.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
a software implementation of the Gating-ML Candidate Recommendation
(CR), an emerging ISAC standard developed in collaboration with the
ISAC Data Standards Task Force, which makes it possible to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

\subsection*{Standard flow operations}
Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, most often in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. Typically, the results of these operations are
again \Rclass{flowFrame} or \Rclass{flowSet} objects. While
transformation, and to a certain extent compensation, are fairly
routine operations with only limited potential for improvement, being
able to implement new methodologies for gating of FCM data, and
extend the capabilities of \Rpackage{flowCore} through object
oriented programming are features that clearly sets our framework
apart from other FCM analysis tools. By factoring out as much of the
bookkeeping as possible, programmers can focus on the actual
operations rather then having to deal with the tedious details of data
integration and access. Third-party methods can act on their own as
first-class citizens in the analysis framework, without breaking the
work flow or the basic infrastructure. This design allows for the
straightforward extension of \Rpackage{flowCore}'s capabilities, and
has already fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

@@

\subsection*{Related flow packages}
In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. As exemplified in the previous section, the
\Rpackage{flowViz} package \cite{sarkar2008ufv} provides sophisticated
data visualization tools, that make use off multivariate trellis
plotting \cite{lattice}.  These functions can be used to quickly
generate customized plots for extended cytometry data sets for both
direct data inspection and quality control.  The objects metadata
information can be used to arrange the layout and composition of the
plots.  Furthermore, the design and the API of the visualization
software is very generic, and users can readily extend its
capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow cytometry
data analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
Project name: flowCore ;
Project home page: http://bioconductor.org ;
Operating system(s): The flowCore package can be installed on all the 
platforms supporting R. These include a wide variety of UNIX platforms, Windows and MacOS.;  
Programming language: R ;
Licence: The Artistic License, Version 2.0 .

@@

    \subsection*{Figure 3 - Quality assessment}
    \label{xyplot}{
        Scatterplot matrix of a single \Rclass{flowSet} from an experiment
        focusing on immune tolerance following kidney
        transplantation. Outlines of the gating regions identified by an
        automated gating operation are added on top of the density
        representation of the data.}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Data transformations}
    \begin{table}[ht]
      \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
        corresponding to value being transformed, $y$ is the result of the
        transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
        constants affecting the transformation function, $e$ is the base of
        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
      \begin{center}
        \begin{tabular}{|l|l|}
          \hline
          \multicolumn{2}{|c|}{Data Transformations} \\
          \hline
          linear & $ax + b$ \\
          quadratic & $ax^2 + bx + c$ \\
          natural logarithm & $ln(x)(r/d)$ \\
          logarithm & $log(x)(r/d)$ \\
          biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
          logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
          truncate & $x_{x<=a} = a$ \\
          scale & $(x-a)/(b-a)$ \\
          split-scale & combination of linear and logarithm \\
          arcsinh & $asinh(a + bx)+c$ \\
          \hline
        \end{tabular}
      \end{center}
    \end{table}

>>>>

\title{flowCore: a Bioconductor package for high throughput flow cytometry}

@@

%% I think switching the last 2 paragraphs make more sense as the 2nd
%% one of this section talk about other closed-source soft etc.
%% But up to you
\Rpackage{flowCore} does not provide a graphical user interface and
all operations are done using a command line interface.  It is
possible to add a more elaborate user interface on top of this
infrastructure, however the focus in this paper is on a programmatic
approach to enable the convenient development of novel analysis
methods and automation of complex analysis approaches.  By taking the
burden of data management from the programmer, and by providing
well-defined application programming interfaces (APIs), it is possible
to readily test new ideas and to easily extend the framework's
functionality.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
a software implementation of the Gating-Markup Language Candidate
Recommendation (Gating-ML CR), an emerging standard developed in collaboration
with the International Society for Analytical Cytology (ISAC) Data
Standards Task Force, which makes it possible to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

\subsection*{Standard flow operations}
Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, most often in an
interactive, graphical user interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. Typically, the results of these operations are
again \Rclass{flowFrame} or \Rclass{flowSet} objects. While
transformation, and to a certain extent compensation, are fairly
routine operations with only limited potential for improvement, being
able to implement new methodologies for gating of FCM data, and extend
the capabilities of \Rpackage{flowCore} through object oriented
programming are features that clearly sets our framework apart from
other FCM analysis tools. By factoring out as much of the bookkeeping
as possible, programmers can focus on the actual operations rather
then having to deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

@@

\subsection*{Related flow packages}
In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. As exemplified in the previous section, the
\Rpackage{flowViz} package \cite{sarkar2008ufv} provides sophisticated
data visualization tools, that make use off multivariate trellis
plotting \cite{lattice}.  These functions can be used to quickly
generate customized plots for extended cytometry data sets for both
direct data inspection and quality control.  The objects metadata
information can be used to arrange the layout and composition of the
plots.  Furthermore, the design and the API of the visualization
software is very generic, and users can readily extend its
capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance
methodology and a framework to create interactive web-based reports of
quality assurance results. The \Rpackage{flowUtil} package implements
data import and export including flow-cytometry specific standard
markup language. Finally, the \Rpackage{flowStats} package provides
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
Project name: flowCore ;
Project home page: http://bioconductor.org ;
Operating system(s): A wide variety of UNIX platforms, Windows and MacOS.;  
Programming language: R ;
Licence: The Artistic License, Version 2.0 .

The flowCore package and its associated packages are part of the
R/Bioconductor project, an environment for statistical computing and
bioinformatics. The R software environment is freely available at
http://www.r-project.org. flowCore and its dependencies (flowQ,
flowViz) are available on the Bioconductor project website
(http://bioconductor.org) as freely distributed and open source
software packages with an Artistic license. They are fully integrated
into the R/Bioconductor environment for statistical computing and
bioinformatics and run on operating systems Windows, Mac OS X, and
Unix.

@@

    \subsection*{Figure 3 - Batch gating}
    \label{xyplot}{ 
      Scatterplot matrix of a single \Rclass{flowSet}
      from an experiment focusing on immune tolerance following kidney
      transplantation. Outlines of the gating regions identified by a
      \Rclass{curve2Filter} automated gating operation are added on
      top of the density representation of the data.}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Data transformations}
    \begin{table}[ht]
      \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
        corresponding to value being transformed, $y$ is the result of the
        transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
        constants affecting the transformation function, $e$ is the base of
        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
      \begin{center}
        \begin{tabular}{|l|l|}
          \hline
          \multicolumn{2}{|c|}{Data Transformations} \\
          \hline
          linear & $ax + b$ \\
          quadratic & $ax^2 + bx + c$ \\
          natural logarithm & $log_e(x)(r/d)$ \\
          logarithm & $log_b(x)(r/d)$ \\
          biexponential & $root(ae^{(b*y)}-ce^{(-d+y)}+f)$ \\
          logicle& $root(Te^{-(m-w)}(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
          truncate & $x_{x{\leq}a} = a$ \\
          scale & $(x-a)/(b-a)$ \\
          split-scale & combination of linear and logarithm \\
          arcsinh & $arcsinh(a + bx)+c$ \\
          \hline
        \end{tabular}
      \end{center}
    \end{table}
